Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-12-03 paper.Rnw
jp comment 22-23
- took out ANOVA estimate
- added summary of results, interpreted (lack of) differences
- clarified HC standard errors
- left in 0s for rho and Tz, but explained why they're there

Our main parameter of interest here is $p_0$, the proportion of
students who are low-hint users.
We estimated $p_0\approx$\Sexpr{prettyNum(round(1-mixProp,2))},
classifying \Sexpr{round(mixProp*100)}\% of students as
high-hint users, so we chose $c$ as the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
This definition agreed with the model's classification (based on
$\eta$ as opposed to $\bar{h}$ in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

How does requesting hints within CTA1 affect learning?
More precisely, would students have achieved higher posttest scores
had they requested hints more (or less) often?

@@

% The classic approach to estimating effects from blocked randomization
% designs is a fixed-effects regression \citep{fisher1926arrangement},
% regressing outcomes on an indicator for treatment assignment and a set
% of indicators for matched sets.
% The results from this analysis are displayed on the row labeled
% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
% errors and 95\% confidence intervals.
% They suggest that a greater-than-median proclivity to request hints
% \emph{hurts} student learning by a large amount, roughly equal to the
% entire treatment effect of CTA1 estimated in
% \citet{pane2014effectiveness}.
Table \ref{tab:matchResults} gives three treatment effect estimates
under matched ignorability.
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with a natural spline with five degrees of freedom for pretest (the standard error
and confidence interval using the ``HC3'' sandwich estimator,
\citealt{sandwichPackage}).
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
the others.

@@

matchTEtab <-
  cbind(
    data.frame(Effect=c('ANOVA','ATE','TOT','ANCOVA')),
    rbind(anovat,atet,tott,ancovat)[,c('Estimate','Std. Error')],
    CI=vapply(list(anovaci,ateci,totci,ancovaci),printci,'a'),
    `$T_Z$`=0,
    `$\\rho^2$`=0
  )

@@

Matched ignorability is a very strong assumption, so these results
should not be taken at face value---especially if weaker students ask for hints at
higher rates, and also score lower on the posttest.
While adjusting for pre-test scores via matching and/or regression
may account for some of that bias, pre-test may not capture all of the
relevant pre-treatment information.
To account for the possibility of unmeasured confounding, we conducted
a sensitivity analysis following the method of \citet{hhh}.
This method imagines a hypothetical missing covariate $U$ and estimates
how the the omission of $U$ alters the estimate
and its standard error.
The result is a ``sensitivity interval'' \citep[c.f.][]{rosenbaum2002observational} that,
with 95\% confidence, contains the true effect accounting for both
sampling uncertainty and uncertainty due to possible confounding.
$U$ is characterized by two sensitivity parameters:
$\rho^2$ is the squared partial correlation between $U$ and $Y$, after
accounting for the observed covariates in the model, and $T_Z$ is the
t-statistic from an ordinary least squares regression of $H$ on $U$, along with the observed covariates.
These sensitivity parameters may be benchmarked by estimating their
counterparts when observed covariates are left out of the model.
The row of Table \ref{tab:matchResults} labeled ``[Pretest]''
imagines a $U$ that predicts $Y$ and $H$ to a similar extent as
pretest.
The resulting sensitivity interval is quite wide, including
unbelievably large negative effects as well as substantial positive
effects.
A slightly less conservative sensitivity analysis, labeled
``[Pretest/2]'' allows $U$ to
predict $H$ and $\rho$ half as well as pretest scores.
This sensitivity interval is wide, but entirely negative.

@@

Our solution was to dichotomize hint requests, as
described above, essentially re-defining the causal question.
\Sexpr{round((mixProp)*100)}\% of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the
\Sexpr{round((1-mixProp)*100)}th percentile of the treatment group.
To make sense of the respective roles of $H$ and $\bar{h}$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\bar{h}$.
In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\bar{h}$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

>>>>

Our main parameter of interest here is $p_0$, the proportion of
students who are low hint users.
We estimated $p_0\approx$\Sexpr{prettyNum(round(1-mixProp,2))},
classifying \Sexpr{round(mixProp*100)}\% of students as
high hint users, so we chose $c$ as the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
This definition agreed with the model's classification (based on
$\eta$ as opposed to $\bar{h}$ in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

How does requesting hints within CTA1 affect learning?
More precisely, would low hint users have achieved higher posttest scores
had they requested hints more often?
Would high hint users have achieved higher posttest scores had the
requested hints less often?

@@

% The classic approach to estimating effects from blocked randomization
% designs is a fixed-effects regression \citep{fisher1926arrangement},
% regressing outcomes on an indicator for treatment assignment and a set
% of indicators for matched sets.
% The results from this analysis are displayed on the row labeled
% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
% errors and 95\% confidence intervals.
% They suggest that a greater-than-median proclivity to request hints
% \emph{hurts} student learning by a large amount, roughly equal to the
% entire treatment effect of CTA1 estimated in
% \citet{pane2014effectiveness}.
Table \ref{tab:matchResults} gives three treatment effect estimates
under matched ignorability.
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with a natural spline with five degrees of freedom for pretest.
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}.
All three sets of standard errors
and confidence intervals used the heteroskedasticity-robust ``HC3'' sandwich estimator,
\citealt{sandwichPackage}.

The three estimates all roughly agree that being a high hint user
decreases posttest scores by nearly a quarter of a standard deviation.
That is, requesting hints hurts posttest scores---teachers should
recommend against students' requesting hints.

The similarity of the three estimates boosts our confidence, if only
slightly. The ``ATE'' and ``TOT'' estimates refer to different groups
of students---ATE estimates an average effect for the entire treatment
group, whereas TOT estimates an effect for high hint users. Had these
been very different, we might wonder why the effect of hint requests
varies so much between high and low hint users, and if that suggests
selection bias.
The fact that the additional covariate adjustment of the ANCOVA model
made little difference in the effect estimate suggests that the
propensity score match eliminated all or most of the confounding bias due to
measured covariates; hence, additional adjustment was unnecessary.

@@

matchTEtab <-
  cbind(
    data.frame(Effect=c('ATE','TOT','ANCOVA')),
    rbind(atet,tott,ancovat)[,c('Estimate','Std. Error')],
    CI=vapply(list(ateci,totci,ancovaci),printci,'a'),
    `$T_Z$`=0,
    `$\\rho^2$`=0
  )

stopifnot(min(matchTEtab[,1])<0.2)
stopfinot(max(matchTEtab[,1])>0.25)

@@

None of these estimates addressed the possiblity of confounding from
unmeasured covariates; instead, they relied on matched ignorability.
This would be a particular concern if pretest scores do not completely
capture students' mathematics ability, and if weaker students ask for hints at
higher rates, and also score lower on the posttest.
Then, whatever aspects of student ability are not captures by the
pretest, and correlate with both hint requests and posttest scores,
would confound the estimates in the upper part of Table
\ref{tab:matchResults}.

To account for the possibility of unmeasured confounding, we conducted
a sensitivity analysis following the method of \citet{hhh}.
This method imagines a hypothetical missing covariate $U$ and estimates
how the the omission of $U$ alters the estimate
and its standard error.
The result is a ``sensitivity interval'' \citep[c.f.][]{rosenbaum2002observational} that,
with 95\% confidence, contains the true effect accounting for both
sampling uncertainty and uncertainty due to possible confounding.
$U$ is characterized by two sensitivity parameters:
$\rho^2$ is the squared partial correlation between $U$ and $Y$, after
accounting for the observed covariates in the model, and $T_Z$ is the
t-statistic from an ordinary least squares regression of $H$ on $U$, along with the observed covariates.
These sensitivity parameters may be benchmarked by estimating their
counterparts when observed covariates are left out of the model.
The first three rows of the table have both $\rho^2$ and $T_Z$ equal
to zero, to reflect that they assume no unmeasured confounders.
The row of Table \ref{tab:matchResults} labeled ``[Pretest]''
imagines a $U$ that predicts $Y$ and $H$ to a similar extent as
pretest.
The resulting sensitivity interval is quite wide, including
both large negative effects as well as substantial positive
effects.
A slightly less conservative sensitivity analysis, labeled
``[Pretest/2]'' allows $U$ to
predict $H$ and $\rho$ half as well as pretest scores.
This sensitivity interval is wide, but entirely negative.

@@

Dichotomizing hint requests into $H$, instead of $\bar{h}$, re-defines
the causal question.
To make sense of the respective roles of $H$ and $\bar{h}$, the
continuous measure on which it is based (see Section
\ref{sec:dichotomizingHintUsage}),
we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\bar{h}$.
In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\bar{h}$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.
